\chapter*{Acknowledgments}
\chaptermark{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

I fear that this section will not do justice to the people I have not
the space to thank individually. Alas, here goes. To those whose names
are not specifically listed here, thank you too.

I wish first to thank Shuhei Ono. The work this thesis represents could
not have happened without his bold vision and the license to explore and
question that working in his lab afforded. His unbridled optimism,
breadth of scientific curiosity, personal integrity, and accessibility
to his personnel are traits I one day hope to emulate. I thank him for
his nonjudgmental yet appropriately measured support of many of my
ideas---even those that led nowhere---, for believing in me even when I
found it difficult to do so myself, and for teaching me to change pump
oil, build vacuum lines, and drink Icelandic vodka.

I also wish to thank my thesis committee members Jeff Seewald, Roger
Summons, and John Pohlman, for their input and genuine interest in my
projects, for their coauthorship on one or more manuscripts, and for
serving as sounding boards for questions about life and career.  Meg Tivey is thanked for chairing my defense
and for her comments on this thesis.

Many members of the Ono lab have contributed time or ideas towards the
work in this thesis. I particularly wish to acknowledge Danielle Gruen,
who has been a wonderful friend, stand-in sibling, and multi-talented
colleague. Without her efforts in the field and the lab, we would not
have made the discovery of the ``hot cow'' that is now perhaps the MIT
Hardcore Stable Isotope Laboratory's best-known result. Bill Olszewski
has been a fount of technical wizardry and occasional oddball humor,
both of which are appreciated. Eoghan Reeves, Genming Luo, Andrew
Whitehill, Shikma Zaarur, Yenny Gonzales Ramos, and Jeemin Rhim are
thanked for their company both at and away from the shop.

Other members of the E25-6\textsuperscript{th} floor community have made
it worth coming into work each day. These include my officemates Ana de
Santiago Torio, Gilad Antler, Mirna Daye, and Kristin Woycheese, and
reliable sources of conversation Simone, Sharon, Christine, Xiaolei, Ross, Shane, Emily, Robert, Haitao, Annie, Christopher, Ainara, and Melody. My fellow JP students---including Ning, Nick, Kate, Evan, Rene, and Craig---provided much needed inspiration (and commiseration and distraction) at times. 

I have had the privilege of working in several other labs during the
past six years.  Tanja Bosak served as my co-advisor for my Generals
project on sulfate reducing bacteria. I thank her for welcoming me into
her lab, for helping me become a more succinct writer and shrewd reader, and
for her encouragement at a particularly frustrating point in my
research. I also thank Mak Saito, who took me under his wing my first
year, offered the opportunity to join my first oceanographic cruise (and
first equator-crossing!), and---after I splashed some acid on myself
while hastily preparing reagents before departing for said
cruise---kindly drove me to his house so that I could have a shower, a
change of clothes, and a day-old scone.  I thank Jeff Seewald for making me feel welcome in his group while I
puttered about trying to deuterate his lab.  

Many thanks are due to members of the Bosak, Saito, and Seewald labs. Min Sub Sim
showed me how to work with sulfate-reducing bacteria, and Peter Hedman,
Alex Evans, and Laura Meredith provided assistance at one time or
another during my first two years at MIT. Help or advice from Sean Sylva, Jill McDermott,
Dawn Moran, Matt McIlvin, Tyler Goepfert, and others in the Fye and 
Watson Buildings at WHOI are gratefully acknowledged.

Several other faculty members dispensed much needed advice at one point
or another, including Ed Boyle, Scott Wankel, Dave Glover, Amanda
Spivak, Scott Doney, Carl Lamborg, and Frieder Klein. Bernhard Peucker-Ehrenbrink, Liz Kujawinski,
and Mark Kurz assisted in navigating the administrative waters of the
JP.\@ Rachel Stanley kindly served as chair for my thesis proposal
defense. I thank those instructors in whose courses I did not
actually enroll, but who still allowed me to participate in their study
tours---David McGee, Taylor Perron, and Oli Jagoutz (MIT),
Rob Sohn, Meg Tivey \& Andrew Daly (Geodynamics 2015, WHOI), John Shaw
(Harvard), and Grant Garven (Tufts).

The administrators in E25, particularly Melody Abedinejad, Mary Eliff,
and (briefly) Annora Borden, are thanked for making life easier in too
many ways to name. I also thank the staff of the MIT Joint Program
Office (Ronni Schwartz and Kris Kipp), PAOC administration (Christine
Maglio), ERL administration (Anna Shaughnessy and Josh Kastorf), EAPS HQ
(esp. Roberta Allard, Angela Ellis, Brian Smith, and Mark Pendleton),
the WHOI MC\&G Department (esp.\ Mary Zawoysky), and the Academic
Programs Office (Julia Westwater, Lea Fraser, Tricia Morin Gebbie,
Christine Charette, Valerie Caron, Meg Tivey, and Jim Yoder). Also, I
want to acknowledge the security staff at WHOI---they more than once let
me into a locked building or gave me a lift back to Winding Lane---, as
well as the custodial staff at MIT, particularly Sergio Medina, who I
could always count on to share stories and snacks when working late into
the night.

I have been lucky enough to be plugged into several wonderful scientific
communities outside of MIT and WHOI, thanks in no small part to Shuhei
and several collaborators. Thanks goes to the instructors and students
of the International Geobiology Summer Course 2013 (it didn't suck!) for
exposing me to the geobiological sciences, and the Agouron Institute and
the Henry Houghton fund of PAOC for sponsoring my participation. The
Deep Carbon Observatory (DCO) funded much of the work in this thesis,
and also afforded me the opportunity to attend and present at several
meetings; Craig Schiffries, Mitch Sogin, Barbara Sherwood Lollar,
Kai-Uwe Hinrichs, and Fumio Inagaki, among others, are thanked for these
opportunities. Two other much-needed breaks punctuated my graduate
studies. I thank Jen Shosa, Michael Lawson, Mike Formolo, Glenn
Hieshima, and everyone else who made being an intern at ExxonMobil
interesting and fun. I also thank my entire IODP Expedition 370
``T-Limit'' family, particularly Verena Heuer, Fumio Inagaki, Akira
Ijiri, Flo Schubotz, Hayley Manners, Masanori Kaneko, Bernhard
Viehweger, and Art Spivack for a great few months at sea and in Kochi.

Miscellaneous thanks goes to the folks at Aerodyne Research (Dave
Nelson, Barry McManus, Joanne Shorter, Mark Zahniser, Christoph Dyroff,
and Stanley Huang) for technical support, Issaku Kohl for organizing the
return of a stranded sample, L.~Taras Bryndzia for arranging funding
from Shell through the Society of Energy Fellows at MIT to support my
last year of studies, Chris Clayton for sharing unpublished data which
helped frame my views on the origin of H in thermogenic hydrocarbons,
Carolyn Ruppel for connecting me to the gas hydrates world,
Jeff Dick for technical support related to CHNOSZ,
Jordan Bird for sampling a rice paddy, my previous research mentors 
(Andre Ellis, Oliver Dorigo, and Chintda Santiskulvong) for introducing me to science,
Penny Morrill for telling me something I needed to hear,
and Lisa Morgan Morzel for gifting me her DCO hat.

My friends outside of EAPS and the JP have played no insignificant role
in helping me maintain perspective on life. For this, I thank my
roommates past and present (esp.\ Nate, Diego, and Benedikt), sailing
partner (Pablo), friends from Harvard EPS (Yanpeng and Joe), fellow
Boston-area EEP alumni (including David, Alex, Daniel, and Matt), and
friends from the MIT Concert Band (Melissa, Tom~C., Brynna, Faith, and
many others). Tom Reynolds deserves a special mention; his work ethic
and constant encouragement as conductor has made me a much improved
trumpet player over my almost five years with the band.

Finally, I thank my parents. They have been voices of encouragement and
reason during my happiest days and my darkest. This doctoral thesis
completes what they came to the U.S. decades ago to begin, but
ultimately chose not to finish so that their son might have a better
shot at doing so himself.

\clearpage

\section{Financial support}

The research activities documented in this thesis were made possible by
grants to my advisor from the U.S. National Science Foundation
(NSF award EAR-1250394), the
National Aeronautics and Space Administration (NASA) Astrobiology
Institute (NAI, University of Colorado, Boulder, CAN 7 under
Cooperative Agreement NNA15BB02A), the Department of Energy (DOE, Small Business Innovation Research program, contract DE-SC0004575),
the Alfred P. Sloan Foundation via the Deep Carbon Observatory, and a
Shell Graduate Fellowship through the MIT Energy Initiative.

I completed the bulk of the work in this thesis while being supported by
a National Defense Science and Engineering Graduate (NDSEG) Fellowship
awarded through the Office of Naval Research of the U.S. Department of
Defense. The Stanley W. Watson Fellowship Fund provided support during
my first summer term at WHOI. The Charles M. Vest Presidential
Fellowship at MIT supported me in the first year of my Ph.D. studies. I
received additional support that year through NSF award EAR-1159318
(to S. Ono and T. Bosak) and the Walter \& Adel Hohenstein
Graduate Fellowship of Phi Kappa Phi.  The MIT Earth Resources Laboratory 
and PAOC Houghton Fund funded my attendance at several conferences.