\chapter{Introduction}\label{ch:1}

\input{figures/fig1.1}

Methane is the simplest and most abundant hydrocarbon. \mrefs[]{Figure}{fig:1:1} shows some statistics on the
portions of the methane cycle that this thesis touches upon.

\section{Essential definitions}\label{sec:1:essential-definitions}

\input{figures/fig1.2}

The \emph{isotopologues} (or isotopic homologues) of a compound have the
same elemental composition and chemical structure, but differ only in
the identity of the isotopes of one or more atoms. The word
\emph{isotopologue} is also seen as \emph{isotopolog}.

\input{figures/fig1.3}

\input{figures/fig1.NA}

The goal of this thesis is to help map the distribution and behavior of
the four most abundant methane isotopologues at the Earth surface and in
the crustal subsurface. These isotopologues are shown in \mrefs[]{Figs.}{fig:1:2} and~\ref{fig:1:3} (approximate ranges of abundance for the singly-substituted methane isotopologues are 
shown in \autoref{fig:1:NA}) and
written in the isotope exchange reaction below:
\begin{equation}\label{eqn:1:exchange}
{}^{13}\text{CH}_4+ {}^{12}{\text{CH}}_3\text{D}\rightleftharpoons {}^{13}{\text{CH}}_3\text{D}+ {}^{12}{\text{CH}}_4
\end{equation}

\input{figures/fig1.4}

In this reaction, one deuterium (D) is exchanged for one hydrogen (H),
while leaving in place the two different carbon (C) isotopes and the
three other H's to which each C is connected. The equilibrium constant
for this reaction is primarily a function of temperature (the
effect of pressure is negligible at near-surface conditions), and is
shown in \autoref{fig:1:3}. The equilibrium constant for this reaction
asymptotically approaches unity as temperatures increase towards
infinity. A sample of methane whose relative abundance of isotopologues
obey the relation
\cee{(\textsuperscript{13}CH\textsubscript{4})(\textsuperscript{12}CH\textsubscript{3}D)
=
(\textsuperscript{13}CH\textsubscript{3}D)(\textsuperscript{12}CH\textsubscript{4})}
(i.e., has a reaction quotient equal to unity) is said to have a
\emph{stochastic distribution} of isotopes among isotopologues. At
lower temperatures, the equilibrium constant is greater than one, albeit
only slightly, by 0.6\% or 6‰ (permil) at room temperature. The origin
of this \emph{clumpiness} at equilibrium at lower temperatures arises from a disproportionate lowering of zero-point energy upon clumping of two or more heavy isotopes (\autoref{fig:1:4}). For more on this topic, readers are referred to \textcite{Eiler_2007_EPSL}.

\input{figures/fig1.5}

Attainment of equilibrium in CH\textsubscript{4} clumped isotopologue
abundances requires reordering of the C--H bonds within molecules. This
may occur by homogeneous (direct) exchange of H between two
CH\textsubscript{4} molecules, or by all CH\textsubscript{4} molecules
independently exchanging H with a second species (heterogeneous).
Understanding the mechanisms enabling exchange in various environments
is vital for correct interpretation of classical and novel stable
isotope geothermometers. \mrefs[]{Figure}{fig:1:5} shows the many pathways by which
several single-carbon compounds can exchange isotopes with compounds in
the C--O--H system.

\section{Preview of thesis content}\label{preview-of-thesis-content}

Several labs are now able to make measurements of the reaction quotient
of \mrefs{Reaction}{eqn:1:exchange}, to better than 0.05\% (or 0.5‰). These include John
Eiler's lab at Caltech \parencite{Stolper++_2014_GCA}, Shuhei Ono's lab at MIT
\parencite{Ono++_2014_AC}, and Ed Young's lab at UCLA \parencite{Young++_2016_IJMS}. For
those interested in the race towards measuring intact methane
isotopologues, readers are referred to \textcite{Jones_2012_Earth}.

\mrefs[]{Chapters}{ch:2}, \ref{ch:3}, and \ref{ch:4} and \autoref{dx:A} describe insights we have gleaned
while studying the origin of C, H, and carbon-hydrogen bonds in
CH\textsubscript{4} using measurements and models of the abundance and
behavior of methane isotopologues. \autoref{ch:2} presents the first survey
of the abundance of fully-resolved
\textsuperscript{13}CH\textsubscript{3}D in various environments on
Earth, and shows how and why microbial methanogenesis occurring under
high H\textsubscript{2} and low CO\textsubscript{2} levels might leave a
very distinct record in the isotopic composition of the
CH\textsubscript{4} produced (\autoref{fig:1:6}). \autoref{ch:2} also briefly touches on
a potential for hydrogen exchange observed in high-maturity thermogenic
gases.

\input{figures/fig1.6}
\input{figures/fig1.7}

\autoref{ch:3} presents a study that attempts to address the oft-contentious
question of where and how methane in seafloor hydrothermal systems
forms. A diagram showing the several main proposed avenues for methane
formation in such systems is in \autoref{fig:1:7}. There is interest in knowing the
answer to this question because of potential implications for the origin
of life at deep-sea hot springs.

\autoref{ch:4} is an experimental and theoretical study that illustrates how
one major sink of methane in the environment, aerobic methanotrophy,
affects the isotopologue abundances of leftover methane (\autoref{fig:1:8}). The
results and equations can be generalized to other major methane sinks \parencite[including oxidation by OH and Cl in the atmosphere;][]{Whitehill++_2017_GCA}, and to other isotopologues (e.g.,
\textsuperscript{12}CH\textsubscript{2}D\textsubscript{2}).

\input{figures/fig1.8}
\input{figures/fig1.9}

Research on the behavior of methane isotopologues like
\textsuperscript{13}CH\textsubscript{3}D have a natural alignment to
many of the questions that are important and possibly unanswered in
assessment of petroleum systems, particularly in poorly understood
basins (\autoref{fig:1:9}). In particular, measurements of methane isotopologues can
\begin{itemize}
	\item
	place quantitative constraints on the stability and origin of C--H
	bonds in hydrocarbons in the Earth's subsurface;
	\item
	test interpretive models of natural gas composition and gas isotope
	systematics; and
	\item
	be used to help anchor the chemistry of natural gases to time and
	temperature.
\end{itemize}

\mrefs[]{Appendices}{dx:A} and \ref{dx:B} represent some initial efforts to read the
hydrogen-isotope and clumped isotopologue record of natural gases and
define what those records mean.

\autoref{dx:C} offers miscellaneous tricks, tips, and data that didn't fit
anywhere else.

























