\chapter{Summary and Outlook}\label{ch:5}

\input{figures/fig5.1}

The preceding chapters presented a set of data (\autoref{fig:5:1}) that shows
several insights into the origin and fate of the methane stable
isotopologues in the environment. As of the time of writing, methane
clumped isotoplogue data have appeared in at least a dozen articles \parencite{Douglas++_2016_GCA,Stolper++_2014_GCA,Stolper++_2014_S,Stolper++_2015_GCA,Wang++_2015_S,Whitehill++_2017_GCA,Ono++_2014_AC,Wang++_2016_GCA,Lopes++_2016_JoDS,Inagaki++_2015_S,Young++_2016_IJMS,Young++_2017_GCA}.

\input{figures/fig5.2}

The Δ\textsuperscript{13}CH\textsubscript{3}D data were shown to be
independent of and complementary to δ\textsuperscript{13}C and δD. A
case was built for why hydrogen (or free energy) exerts a major control
on Δ\textsuperscript{13}CH\textsubscript{3}D and D/H of methane; this
and several other controls are shown in \autoref{fig:5:2}. And several
opportunities for advancement on the technical and theoretical sides of
the problem of methane origin have been highlighted (\autoref{fig:5:3}), including:
\begin{itemize}
	\item
	refining the calibration at low temperatures \parencite{Larson+Hall_1965_JPC,Naito++_2005_ApCat,Golden++_2001_JACS,Robertson++_1975_TFS};
	\item
	experiments to retrieve kinetics associated with the breaking and
	reforming of C--H bonds \parencite{Reeves++_2012_GCA,Lyon+Hulston_1984_GCA,Koepp_1978};
	\item
	construction of numerical models to test hypotheses regarding
	biophysical controls on isotopologue abundances.
\end{itemize}

\input{figures/fig5.3}

Through this work, we have also realized that there is much more
information to be extracted from simple gas chemistry and conventional
stable isotope data than has perhaps been widely appreciated; or that
has been somewhat overlooked in the rush to develop and deploy new
technologies. Measurements of clumped isotopologues provide more
dimensions of data to characterize methane samples, which, as this
thesis shows, does provide additional constraints on certain Earth
systems. However, instruments to measure the doubly-substituted
isotopologues will likely remain a novelty for the foreseeable future.

I suspect that the most significant advances for our understanding of
the origins, transport, and fate of methane are still yet to come. The
biggest steps forward will come with moving away from straightforward
but often restrictive phenomenological representations of legacy data
and towards predictive and quantitative approaches to testing of
well-defined and geologically-plausible hypotheses. Isotopologue data
will certainly play a large role in helping narrow the solution space of
problems of geochemical nature. \mrefs[]{Figures}{fig:5:4} and \ref{fig:5:5} show some attempts at
making such data more easily accessible.

\input{figures/fig5.4}
\input{figures/fig5.5}




